---
title: "Manhattan Real Estate Analysis"
author: "Hansen Han"
date: "2024-03-08"
output:
  html_document: default
  pdf_document: default
---

## Load Packages

```{r errors=FALSE, messages=FALSE}
library(tidyverse)
library(magrittr)
library(ggpubr)
library(FinancialMath)
library(ggsignif)
```

## Loading Data & Pre-Processing

Neighborhood - Zipcode Matching Data: We'll need this to assign each property to a neighborhood

```{r}
neighborhood_data <- read_csv("data/nyc_zip_borough_neighborhoods_pop.csv")
neighborhood_data
```

Rental Data:

```{r}
rental_data <- read_csv("data/manhattan-ny_rent_2024_03_15.csv")
rental_data %<>% left_join(neighborhood_data, by=c("unit_zipcode"="zip"))  %>% filter(borough == "Manhattan") # attach neighborhood data
rental_data
```

Sales Data:

```{r}
sales_data <- read_csv("data/manhattan-ny_sale_2024_03_15.csv")
#rename some variables
sales_data$price <- sales_data$unformattedPrice
sales_data$sqft <- sales_data$area
sales_data %<>% left_join(neighborhood_data, by=c("addressZipcode"="zip")) %>% filter(borough == "Manhattan")
sales_data
```

Annotated Sales Data: (this is data where which we have more than 2 historical sales) 1583/6992 = 0.22, so we have about 22% of the sales properties actively listed

```{r errors=FALSE, messages=FALSE}
sales_data_annotated <- read_csv("data/manhattan-ny_sale_2024_03_15_with_cagr_and_hoa_fees.csv")
#rename some variables
sales_data_annotated$price <- sales_data_annotated$unformattedPrice
sales_data_annotated$sqft <- sales_data_annotated$area
sales_data_annotated %<>% left_join(neighborhood_data, by=c("addressZipcode"="zip"))  %>% filter(borough == "Manhattan")
sales_data_annotated
```

Add financing information (assuming a 20% downpayment on a 30 year mortgage)
```{r}
interest_rate = 0.0772
downpayment_percent = 0.2
air12 = interest_rate / 12 #annual interest rate/12
installments = 12 * 30# assuming a 30-year fixed 

sales_data_annotated %<>% mutate(downpayment = unformattedPrice*downpayment_percent) %>% mutate(monthly_payment = (unformattedPrice-downpayment)*(air12*(1+air12)^installments)/((1+air12)^installments - 1))

sales_data_annotated %<>% mutate(monthly_payment_w_fees = monthly_payment + hoa_fee)
sales_data_annotated
```

Add CONDO vs CO-OP
If you look up online, "D4-ELEVATOR APT", "C6-WALK-UP APARTMENT" buildings are CO-OP buildings while R0-CONDOMINIUMS are CONDO buildings. There are some other weird types in rare cases but they are very few and we want to ignore them for this analysis. 
```{r}
sales_data_annotated %<>% filter(building_type %in% c("D4-ELEVATOR APT", "C6-WALK-UP APARTMENT", "R0-CONDOMINIUMS")) %>% mutate(building_type_simple = case_when(
  building_type %in% c("D4-ELEVATOR APT", "C6-WALK-UP APARTMENT	") ~ "CO-OP",
  building_type %in% c("R0-CONDOMINIUMS") ~ "CONDO"
))

sales_data_annotated %<>% filter(building_type_simple %in% c("CO-OP", "CONDO"))
sales_data_annotated # 1583 - 1368  = 215 (listings removed)
```
## Analysis

### Data Exploration

#### Modeling Sales / Rental Price
I'm not really interested in this, so this won't go very deep, but I thought it would be an interesting to give it a shoot and see what comes out...


How beds, bathrooms, sqft, condo vs. co-op and neighborhood affect sales price
```{r}
lm_fit <- lm(formula = price ~ beds + baths + neighborhood + sqft + building_type_simple, data = sales_data_annotated)
summary(lm_fit)
```
0.59 R^2....its something, but its not great! 
Interesting baths is more significant than beds (which I'm certain are correlated, but baths probably conveys more information)

Now lets look at rental data
```{r}
rental_data$sqft <- rental_data$area
lm_fit <- lm(formula = price ~ beds + baths + neighborhood + sqft, data = rental_data)
summary(lm_fit)
```

Wow, 0.76 R^2, that's not bad!

We actually have in rental data the "zestimate" which is Zillow's predictions, lets see how correlated this compares against that 
```{r}
predicted_values <- predict(lm_fit, newdata = sales_data)
sales_data$predicted_rent <- predicted_values
sales_data %>% select(zestimate, predicted_rent) %>% drop_na() %>% cor()
```

0.55, so not that great! Zillow's estimate is pretty different. I saw some negative values in there too, which is a clear sign of not being in the same range and bad extrapolation! 

#### Listing Distribution

```{r}
# Tally neighborhoods
neighborhood_counts_rentals <- data.frame(table(rental_data$neighborhood))
neighborhood_counts_sales <- data.frame(table(sales_data$neighborhood))
```


Is the subset of data generally representative of the total sales data (is it skewed towards some neighborhoods over others)?
```{r}
neighborhood_counts_sales_annotated <- data.frame(table(sales_data_annotated$neighborhood))
neighborhood_counts_sales_annotated %>% arrange(desc(Freq))
```

Compare the ratio of frequency for each neighborhood, in a perfectly representative subset, the ratio would be exactly the same for each. Here they're generally between 4-6. I would say the general order is relatively the same and the ratio is pretty consistent, so I think its generally well sampled in terms of neighborhoods. 

```{r}
neighborhood_counts_sales %>% left_join(neighborhood_counts_sales_annotated, by = c("Var1"="Var1")) %>% mutate(ratio = `Freq.x`/`Freq.y`)
```

Compare ratio of rental to sales properties by neighborhood
```{r}
neighborhood_counts_rentals$num_rental_properties <- neighborhood_counts_rentals$Freq
neighborhood_counts_rentals$Freq <- NULL

neighborhood_counts_sales$num_sales_properties <- neighborhood_counts_sales$Freq
neighborhood_counts_sales$Freq <- NULL

neighborhood_counts <- neighborhood_counts_rentals %>% left_join(neighborhood_counts_sales, by = c("Var1"="Var1"))
neighborhood_counts %<>% mutate(ratio = num_rental_properties/num_sales_properties)
neighborhood_counts
```


### Renting vs. Buying: Monthly Cost
Assuming that the buyer is taking out an 80% mortgage, what is the monthly cost of owning an apartment vs. renting one in Manhattan? 

#### All Neighborhoods
Interestingly rent is higher for studios, then otherwise owning is more expensive.
```{r message=FALSE, warning=FALSE}
sales_data_annotated %>% mutate(type = "sale") %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) %>% ggplot(aes(x = type, y = monthly_payment, color = type)) + geom_boxplot(outlier.shape =NA) + geom_jitter(alpha=0.2, size=0.3) + theme_bw() + xlab("Type") + ylab("Monthly Payment ($)") + theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Monthly Payment in Rentals vs. Sales") + scale_y_log10() + facet_wrap(~beds) +
  geom_signif(comparisons = list(c("rental", "sale")), 
              map_signif_level = TRUE, 
              textsize = 3, 
              tip_length = 0.02, 
              step_increase = 0.1,
              y_position = 3)
```

#### By Neighborhood
Although interesting overall, neighborhood also plays apart, so we should look at that:

##### Studios
```{r warning = FALSE}
sales_data_annotated %>% mutate(type = "sale") %>% filter(beds == 0) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% filter(beds == 0) %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) %>% ggplot(aes(x = type, y = monthly_payment, color = type)) + geom_boxplot(outlier.shape =NA) + geom_jitter(alpha=0.2, size=0.3) + theme_bw() + xlab("Type") + ylab("Monthly Payment ($)") + theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Monthly Payment in Rentals vs. Sales for Studio Apartments") + scale_y_log10() + facet_wrap(~neighborhood) +
  geom_signif(comparisons = list(c("rental", "sale")), 
              map_signif_level = TRUE, 
              textsize = 3, 
              tip_length = 0.02, 
              step_increase = 0.1,
              y_position = 4)


sales_data_annotated %>% mutate(type = case_when(
  building_type_simple == "CONDO" ~ "sale_condo",
  building_type_simple == "CO-OP" ~ "sale_coop"
)) %>% filter(beds == 0) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% filter(beds == 0) %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) %>% ggplot(aes(x = type, y = monthly_payment, color = type)) + geom_boxplot(outlier.shape =NA) + geom_jitter(alpha=0.2, size=0.3) + theme_bw() + xlab("Type") + ylab("Monthly Payment ($)") + theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Monthly Payment in Rentals vs. Sales for Studio Apartments") + scale_y_log10() + facet_wrap(~neighborhood) +
  geom_signif(comparisons = list(c("rental", "sale_condo"), c("rental", "sale_coop")), 
              map_signif_level = TRUE, 
              textsize = 3, 
              tip_length = 0.02, 
              step_increase = 0.1,
              y_position = 3)

```
##### 1 Bedroom
```{r warning = FALSE}
sales_data_annotated %>% mutate(type = "sale") %>% filter(beds == 1) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% filter(beds == 1) %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) %>% ggplot(aes(x = type, y = monthly_payment, color = type)) + geom_boxplot(outlier.shape =NA) + geom_jitter(alpha=0.2, size=0.3) + theme_bw() + xlab("Type") + ylab("Monthly Payment ($)") + theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Monthly Payment in Rentals vs. Sales for 1 Bedroom Apartments") + scale_y_log10() + facet_wrap(~neighborhood) +
  geom_signif(comparisons = list(c("rental", "sale")), 
              map_signif_level = TRUE, 
              textsize = 3, 
              tip_length = 0.02, 
              step_increase = 0.1,
              y_position = 4)


sales_data_annotated %>% mutate(type = case_when(
  building_type_simple == "CONDO" ~ "sale_condo",
  building_type_simple == "CO-OP" ~ "sale_coop"
)) %>% filter(beds == 1) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% filter(beds == 1) %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) %>% ggplot(aes(x = type, y = monthly_payment, color = type)) + geom_boxplot(outlier.shape =NA) + geom_jitter(alpha=0.2, size=0.3) + theme_bw() + xlab("Type") + ylab("Monthly Payment ($)") + theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Monthly Payment in Rentals vs. Sales for 1 Bedroom Apartments") + scale_y_log10() + facet_wrap(~neighborhood) +
  geom_signif(comparisons = list(c("rental", "sale_condo"), c("rental", "sale_coop")), 
              map_signif_level = TRUE, 
              textsize = 3, 
              tip_length = 0.02, 
              step_increase = 0.1,
              y_position = 3)

```
##### 2 Bedroom

```{r warning = FALSE}
sales_data_annotated %>% mutate(type = "sale") %>% filter(beds == 2) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% filter(beds == 2) %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) %>% ggplot(aes(x = type, y = monthly_payment, color = type)) + geom_boxplot(outlier.shape =NA) + geom_jitter(alpha=0.2, size=0.3) + theme_bw() + xlab("Type") + ylab("Monthly Payment ($)") + theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Monthly Payment in Rentals vs. Sales for 2 Bedroom Apartments") + scale_y_log10() + facet_wrap(~neighborhood) +
  geom_signif(comparisons = list(c("rental", "sale")), 
              map_signif_level = TRUE, 
              textsize = 3, 
              tip_length = 0.02, 
              step_increase = 0.1,
              y_position = 4)


sales_data_annotated %>% mutate(type = case_when(
  building_type_simple == "CONDO" ~ "sale_condo",
  building_type_simple == "CO-OP" ~ "sale_coop"
)) %>% filter(beds == 2) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% filter(beds == 2) %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) %>% ggplot(aes(x = type, y = monthly_payment, color = type)) + geom_boxplot(outlier.shape =NA) + geom_jitter(alpha=0.2, size=0.3) + theme_bw() + xlab("Type") + ylab("Monthly Payment ($)") + theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Monthly Payment in Rentals vs. Sales for 2 Bedroom Apartments") + scale_y_log10() + facet_wrap(~neighborhood) +
  geom_signif(comparisons = list(c("rental", "sale_condo"), c("rental", "sale_coop")), 
              map_signif_level = TRUE, 
              textsize = 3, 
              tip_length = 0.02, 
              step_increase = 0.1,
              y_position = 3)
```

##### 3 Bedroom

```{r warning=FALSE}
sales_data_annotated %>% mutate(type = "sale") %>% filter(beds == 3) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% filter(beds == 3) %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) %>% ggplot(aes(x = type, y = monthly_payment, color = type)) + geom_boxplot(outlier.shape =NA) + geom_jitter(alpha=0.2, size=0.3) + theme_bw() + xlab("Type") + ylab("Monthly Payment ($)") + theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Monthly Payment in Rentals vs. Sales for 3 Bedroom Apartments") + scale_y_log10() + facet_wrap(~neighborhood) +
  geom_signif(comparisons = list(c("rental", "sale")), 
              map_signif_level = TRUE, 
              textsize = 3, 
              tip_length = 0.02, 
              step_increase = 0.1,
              y_position = 4)


sales_data_annotated %>% mutate(type = case_when(
  building_type_simple == "CONDO" ~ "sale_condo",
  building_type_simple == "CO-OP" ~ "sale_coop"
)) %>% filter(beds == 3) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% filter(beds == 3) %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) %>% ggplot(aes(x = type, y = monthly_payment, color = type)) + geom_boxplot(outlier.shape =NA) + geom_jitter(alpha=0.2, size=0.3) + theme_bw() + xlab("Type") + ylab("Monthly Payment ($)") + theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Monthly Payment in Rentals vs. Sales for 3 Bedroom Apartments") + scale_y_log10() + facet_wrap(~neighborhood) +
  geom_signif(comparisons = list(c("rental", "sale_condo"), c("rental", "sale_coop")), 
              map_signif_level = TRUE, 
              textsize = 3, 
              tip_length = 0.02, 
              step_increase = 0.1,
              y_position = 3)
```
##### 4 Bedroom

```{r warning = FALSE}
sales_data_annotated %>% mutate(type = "sale") %>% filter(beds == 4) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% filter(beds == 4) %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) %>% ggplot(aes(x = type, y = monthly_payment, color = type)) + geom_boxplot(outlier.shape =NA) + geom_jitter(alpha=0.2, size=0.3) + theme_bw() + xlab("Type") + ylab("Monthly Payment ($)") + theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Monthly Payment in Rentals vs. Sales for 4 Bedroom Apartments") + scale_y_log10() + facet_wrap(~neighborhood) +
  geom_signif(comparisons = list(c("rental", "sale")), 
              map_signif_level = TRUE, 
              textsize = 3, 
              tip_length = 0.02, 
              step_increase = 0.1,
              y_position = 4)

sales_data_annotated %>% mutate(type = case_when(
  building_type_simple == "CONDO" ~ "sale_condo",
  building_type_simple == "CO-OP" ~ "sale_coop"
)) %>% filter(beds == 4) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% filter(beds == 4) %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) %>% ggplot(aes(x = type, y = monthly_payment, color = type)) + geom_boxplot(outlier.shape =NA) + geom_jitter(alpha=0.2, size=0.3) + theme_bw() + xlab("Type") + ylab("Monthly Payment ($)") + theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Monthly Payment in Rentals vs. Sales for 4 Bedroom Apartments") + scale_y_log10() + facet_wrap(~neighborhood) +
  geom_signif(comparisons = list(c("rental", "sale_condo"), c("rental", "sale_coop")), 
              map_signif_level = TRUE, 
              textsize = 3, 
              tip_length = 0.02, 
              step_increase = 0.1,
              y_position = 3)
```

Generally, it seems that as bedrooms increase, the difference in monthly payment between rentals and sales increase, there also seem to be some strong neighborhood effects. Interestingly, it seems like there isn't a significant difference in overall monthly cost for large swaths of apartments, particularly in the 0 and 1 bedroom ranges. 

#### What % More Expensive is Buying vs. Renting? 

```{r}
test_data <- sales_data_annotated %>% mutate(type = "sale") %>% filter(beds == 0) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% filter(beds == 0) %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) 

test_data %>% filter(sqft >0 ) %>% mutate(monthly_price_sqft = monthly_payment/sqft) %>% filter(type == "sale") %>% pull(monthly_price_sqft) %>% median() / test_data %>% filter(sqft >0 ) %>% mutate(monthly_price_sqft = monthly_payment/sqft) %>% filter(type == "rental") %>% pull(monthly_price_sqft) %>% median()

```

The average monthly cost of a studio in Manhattan is only 5% higher than a Rental (4% median)

```{r}
test_data <- sales_data_annotated %>% mutate(type = "sale") %>% filter(beds == 1) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% filter(beds == 1) %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) 

test_data %>% filter(sqft >0 ) %>% mutate(monthly_price_sqft = monthly_payment/sqft) %>% filter(type == "sale") %>% pull(monthly_price_sqft) %>% median() / test_data %>% filter(sqft >0 ) %>% mutate(monthly_price_sqft = monthly_payment/sqft) %>% filter(type == "rental") %>% pull(monthly_price_sqft) %>% median()
```

10% higher for 1 bedrooms (mean), (8% median)

```{r}
test_data <- sales_data_annotated %>% mutate(type = "sale") %>% filter(beds == 2) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% filter(beds == 2) %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) 

test_data %>% filter(sqft >0 ) %>% mutate(monthly_price_sqft = monthly_payment/sqft) %>% filter(type == "sale") %>% pull(monthly_price_sqft) %>% median() / test_data %>% filter(sqft >0 ) %>% mutate(monthly_price_sqft = monthly_payment/sqft) %>% filter(type == "rental") %>% pull(monthly_price_sqft) %>% median()
```

29% higher for 2 bedrooms (29% median)

```{r}
test_data <- sales_data_annotated %>% mutate(type = "sale") %>% filter(beds == 3) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% filter(beds == 3) %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) 

test_data %>% filter(sqft >0 ) %>% mutate(monthly_price_sqft = monthly_payment/sqft) %>% filter(type == "sale") %>% pull(monthly_price_sqft) %>% median() / test_data %>% filter(sqft >0 ) %>% mutate(monthly_price_sqft = monthly_payment/sqft) %>% filter(type == "rental") %>% pull(monthly_price_sqft) %>% median()
```

57% for 3 bedrooms (47% median)

```{r}
test_data <- sales_data_annotated %>% mutate(type = "sale") %>% filter(beds == 4) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type) %>% full_join(rental_data %>% filter(beds == 4) %>% mutate(monthly_payment = price, type = "rental") %>% filter(!is.na(monthly_payment)) %>% select(neighborhood, beds, baths, sqft, monthly_payment, type)) 

test_data %>% filter(sqft >0 ) %>% mutate(monthly_price_sqft = monthly_payment/sqft) %>% filter(type == "sale") %>% pull(monthly_price_sqft) %>% median() / test_data %>% filter(sqft >0 ) %>% mutate(monthly_price_sqft = monthly_payment/sqft) %>% filter(type == "rental") %>% pull(monthly_price_sqft) %>% median()
```

111% for 4 bedrooms (66% median)

### Buying Condos vs. Co-Ops

#### Are There More Condos or Co-Op Listings?
There seems to be almost double the amount of condos to co-ops listed in this subset
```{r}
building_type_counts <- data.frame(table(sales_data_annotated$building_type_simple))
building_type_counts
```



#### Comparing HOA Fees

Wherever there is data comparing CO-OPs and CONDOs, CO-OPs have much higher HOA fees. 
```{r warning=FALSE}

# Define a custom y position for the p-value labels
custom_label_y <- exp(mean(log(sales_data_annotated$hoa_fee), na.rm = TRUE))

# Assuming `sales_data_annotated` is your data frame
sales_data_annotated %>%
  filter(beds < 6) %>%
  ggplot(aes(x = neighborhood, y = hoa_fee, color = building_type_simple)) + 
  geom_boxplot(outlier.shape = NA) + 
  theme_bw() + 
  xlab("Neighborhood") + 
  ylab("HOA Fees") + 
  theme(legend.position = "right", 
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  ggtitle("HOA Fees by # Beds and Neighborhood") + 
  scale_y_log10() + 
  facet_wrap(~beds) 
```

#### Comparing Price/Sqft

How does the price per squarefoot compare between Co-Ops and Condos? 
```{r}
sales_data_annotated %>% filter(sqft > 0) %>% mutate(monthly_payment_per_sqft = price/sqft) %>% ggplot(aes(x = building_type_simple, y = monthly_payment_per_sqft, color = building_type_simple)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(alpha = 0.2) + 
  theme_bw() + 
  xlab("Neighborhood") + 
  ylab("Price per Sqft") + 
  theme(legend.position = "right", 
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  ggtitle("Price/Sqft of Co-Op vs. Condo") +
  stat_compare_means(method = "t.test", label = "p.signif", comparisons = list(c("CO-OP", "CONDO")))
```
```{r}
price_per_sqft_data <- sales_data_annotated %>%
  filter(sqft > 0) %>%
  mutate(monthly_payment_per_sqft = price / sqft)

# Calculate the mean price per square foot for each building type
mean_price_per_sqft <- price_per_sqft_data %>%
  group_by(building_type_simple) %>%
  summarise(mean_price_per_sqft = mean(monthly_payment_per_sqft, na.rm = TRUE))

mean_price_per_sqft
```
1711.269/1109.424 = 1.54 (about 50% more expensive per sqft on average)

```{r}
sales_data_annotated %>% filter(building_type_simple == "CONDO", neighborhood == "Upper East Side", beds == 1) %>% pull(historical_cagr) %>% median()
```


### Investment Analysis

#### How Do Historic Returns (Broadly) Compare vs. S&P 500?

When I looked it up, I saw the CAGR for the S&P500 is 9.27%, so I'm going to use that as a benchmark to compare against the CAGR for these listings
```{r warning=FALSE}
sales_data_annotated %>% ggplot(aes(x = neighborhood, y = historical_cagr, color = building_type_simple)) + geom_boxplot(outlier.shape =NA) + theme_bw() + xlab("Neighborhood") + ylab("Historical CAGR") + theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("All Units") + scale_y_log10() + geom_hline(yintercept = 0.0927, linetype = "dashed", color = "red")
```
The baseline historical CAGR for these apartments seems to be below the S&P 500 (but you can actually multiply it using a mortgage, which you can't really do with others)

##### Breakdown by Bedroom & Neighborhood

Facet By Beds (1-4)
```{r warning=FALSE}
sales_data_annotated %>% filter(beds < 6) %>% ggplot(aes(x = neighborhood, y = historical_cagr, color = building_type_simple)) + geom_boxplot(outlier.shape =NA)+ theme_bw() + xlab("Neighborhood") + ylab("Historical CAGR") + theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Historic Returns by # Beds and Neighborhood") + scale_y_log10() + geom_hline(yintercept = 0.0927, linetype = "dashed", color = "red") + facet_wrap(~beds)
```


#### Identifying Ideal Property Types
Ideally, we want to identify a neighborhood x apartment type that has high historic returns with lower volatility, and many data points so that we have higher confidence in the data

While 4 bedroom condos in Lower Manhattan have the best overall profile (9% median returns with .2% SD, there are only 2 points so its not reliable)

```{r}

# n = 25 is arbitrary 
sales_data_annotated %>%
  group_by(beds, building_type_simple, neighborhood) %>% filter(n() > 25) %>%
  summarise(
    median_cagr = median(historical_cagr, na.rm = TRUE),
    std_dev = sd(historical_cagr, na.rm = TRUE),
    count = n(),
    ratio = median(historical_cagr, na.rm = TRUE)/sd(historical_cagr, na.rm = TRUE)
  ) %>% arrange(desc(ratio)) 
```

1 Bedroom Condos in the upper east side seem to be the best, with ~5% median CAGR with 4% standard deviation and 39 data points. 


##### Income / Capital Requirements
At the lowest end, you'd need an annual income of at least 137151.5 and $99,0000, but you could own a very good apartment. 
To afford median monthly payment, you'd want $218k annual salary

Minimum:
```{r }
listing_data <- sales_data_annotated

# Use the 40X rule to see what minimum income you'd need to afford
listing_data %>% filter(neighborhood == "Upper East Side", building_type_simple=="CONDO", beds == 1) %>% pull(monthly_payment) %>% min() * 40 # $113,151.5

listing_data %>% filter(neighborhood == "Upper East Side", building_type_simple=="CONDO", beds == 1) %>% pull(downpayment) %>% min() # $99,000

```

Median:
```{r}
# Use the 40X rule to see what minimum income you'd need to afford
listing_data %>% filter(neighborhood == "Upper East Side", building_type_simple=="CONDO", beds == 1) %>% pull(monthly_payment) %>% median() * 40 # $218,302

listing_data %>% filter(neighborhood == "Upper East Side", building_type_simple=="CONDO", beds == 1) %>% pull(downpayment) %>% median()  # $191,000

```


#### 5 Year Projections
We want to see what the equity in the property would be after 5 years, and also what the appreciation on it would be like. To compute this, we use the historical CAGR of the property to project the future value and then calculate the equity built after several years by building an amoritization table for each property. 

```{r}
listing_data$remaining_balance <- NULL
listing_data$equity <- NULL

n_years <- 5 # years 
n_years_in_months <- n_years*12
listing_data$loan_amount <- listing_data$downpayment / downpayment_percent

# Parameters
annual_interest_rate <- 0.0772  # 2% annual interest
compounding_frequency <- 12   # Monthly compounding
payment_frequency <- 12       # Monthly payments
loan_term_years <- 30


balance_at_target_year <- c()
for (i in seq(length(listing_data$`Unnamed: 0`))) {

  # Calculate the amortization table for each property
  amortization_table <- amort.table(
    Loan = listing_data[i, ] %>% pull(loan_amount),
    n = loan_term_years * payment_frequency,  # Total number of payments
    i = annual_interest_rate,
    ic = compounding_frequency,
    pf = payment_frequency,
    plot = FALSE  # Set TRUE if you want to plot payment proportions
  )
  
  # extract the data
  data.frame(amortization_table$Schedule)[n_years_in_months, ] %>% pull(`Principal.Paid`)
  
  balance_at_target_year <- c(balance_at_target_year, data.frame(amortization_table$Schedule)[n_years_in_months, ] %>% pull(`Balance`))
  
    
}

options(scipen = 999)

listing_data$balance_at_target_year <- balance_at_target_year
listing_data$equity_at_target_year <- listing_data$loan_amount - listing_data$balance_at_target_year + listing_data$downpayment
listing_data$balance_paid_off_at_target_year <- listing_data$loan_amount - listing_data$balance_at_target_year 

# this is purely appreciation on the initial down payment (not considering equity)
listing_data$est_roi_on_downpayment <- (listing_data$historical_cagr*n_years*listing_data$unformattedPrice)/listing_data$downpayment
listing_data$est_roi_vs_benchmark <- (listing_data$est_roi_on_downpayment - 0.0927*n_years)/n_years

listing_data
```


##### Comparing Excess Returns
We look at all the listings and their historical CAGR to see how well they would perform against the S&P 500, generally most (if financed) would yield a greater ROI on the down payment, but have high volatility
```{r warning=FALSE}
# Calculate the IQR and determine bounds for outlier removal
listing_data_2 <- listing_data %>% filter(beds < 4) %>%
  group_by(neighborhood, building_type_simple) %>%
  mutate(Q1 = quantile(est_roi_vs_benchmark, 0.25),
         Q3 = quantile(est_roi_vs_benchmark, 0.75)) %>%
  mutate(IQR = Q3 - Q1,
         Lower_Bound = Q1 - 1.5 * IQR,
         Upper_Bound = Q3 + 1.5 * IQR) %>%
  filter(est_roi_vs_benchmark >= Lower_Bound & est_roi_vs_benchmark <= Upper_Bound) %>%
  ungroup()

# only look at neighborhoods with data > 50 (higher certainty)
listing_data_2 %<>% group_by(neighborhood) %>% filter(n() > 50)

# Plotting the filtered data without outliers
ggplot(listing_data_2, aes(x = neighborhood, y = est_roi_vs_benchmark, color = building_type_simple)) +
  geom_boxplot() +  # Now the boxplot will naturally have no outliers as they were filtered out
  geom_jitter(alpha = 0.4) +  # Adds a jitter plot to display all data points
  theme_bw() +  # Applies a minimal theme
  xlab("Neighborhood") +
  ylab("Excess CAGR vs. Benchmark") +
  theme(legend.position = "right", 
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("All Units") + geom_hline(yintercept = 0, linetype = "dashed", color = "red") + facet_wrap(~beds) +
  geom_signif(comparisons = list(c("CO-OP", "CONDO")), 
              map_signif_level = TRUE, 
              textsize = 3, 
              tip_length = 0.02, 
              step_increase = 0.1,
              y_position = 0)

```



#### Sharpe Ratio for our "Ideal" Category
```{r}
bed_num <- 1

portfolio_return <- listing_data %>% filter(beds == bed_num) %>% filter(neighborhood == "Upper East Side") %>% pull(historical_cagr) %>% mean()
risk_free_rate <- 0.035    # current rate for treasury bonds (3.5%) 
standard_deviation <- listing_data %>% filter(beds == bed_num) %>% filter(neighborhood == "Upper East Side") %>% pull(historical_cagr) %>% sd()  # 12%

# Calculate excess return
excess_return <- portfolio_return - risk_free_rate

# Calculate Sharpe Ratio (alpha)
sharpe_ratio <- excess_return / standard_deviation

sharpe_ratio
```

#### What % of Listings Have Higher Historic Returns than S&P 500 (With a Mortgage)
```{r}
listing_data %>% filter(est_roi_vs_benchmark > 0) %>% pull(est_roi_vs_benchmark) %>% length()
```

```{r}
listing_data %>% filter(est_roi_vs_benchmark < 0) %>% pull(est_roi_vs_benchmark) %>% length()
```

After 5 years, 418 did not beat benchmark, 950 did (69% of total)

##### Breakdown by Apt Type

Studio: 71.6%
1 Bedroom: 70.0%
2 Bedroom: 72.4%
3 Bedroom: 61.3%
4 Bedroom: 59.2%
```{r}
bed_num = 4
listing_data %>% filter(beds == bed_num) %>% filter(est_roi_vs_benchmark > 0) %>% pull(est_roi_vs_benchmark) %>% length() / (listing_data %>% filter(beds == bed_num) %>% pull(est_roi_vs_benchmark)) %>% length()

```


#### Are HOA Fees Associated with Returns? 

```{r}
# Try compare hoa fee with cagr directly (no significant relationship)
lm_fit <- lm(formula = historical_cagr ~ hoa_fee, data = listing_data)
summary(lm_fit)
```

```{r}
# Try to normalize by taking hoa fee per sqft (no significant linear relationship)
lm_fit <- lm(formula = historical_cagr ~ hoa_fee_per_sqft, data = listing_data %>% filter(sqft > 0) %>% mutate(hoa_fee_per_sqft = hoa_fee/sqft))
summary(lm_fit)
```
Plot to visually inspect 
```{r warning=FALSE}
listing_data %>% ggplot(aes(x = hoa_fee, y = historical_cagr)) + geom_point() + theme_bw() + scale_y_log10() + scale_x_log10() + geom_smooth()
```



